Asthma Cohort Descriptive Epidemiology

### 2017-12-08 r_scripts/analysis/asthma_cohort_descriptive_epi.Rmd
updated agu figure for kate and updated asthma summary statistics for cohort.

# load outcomes vectors
load("./data/health/outcome_list.RData")
# subset asthma and saba vector
asthma_icd9 <- pluck(outcome_icd9_list, "asthma")
saba_ndc <- pluck(outcome_icd9_list, "saba")

# read in place of service csv file
place_of_service <- read_csv("./data/health/2013-oregon_pos.csv")

@@

```{r cohort_import}
# read path
read_path <- "./data/health/2013-oregon_asthma_cohort.csv"
# read in asthma cohort (reading a million rows for now)
asthma_cohort <- data.table::fread(read_path, 
  # read options
  header = T, nrows = 10000, colClasses = rep("character", 72)) %>% 
  # replace *NULL* with NA
  mutate_all(funs(replace(., .== "*NULL*" | . == "", NA))) %>% 
  # create visit type indicator 
  mutate(visit_type = as.factor(
    case_when((dx1 %in% asthma_icd9) ~ "dx_asthma_primary",
    # asthma but not as a primary
    !(dx1 %in% asthma_icd9) & (dx2 %in% asthma_icd9 | dx3 %in% asthma_icd9 | 
      dx4 %in% asthma_icd9 | dx5 %in% asthma_icd9 | dx6 %in% asthma_icd9 | 
      dx7 %in% asthma_icd9 | dx8 %in% asthma_icd9 | dx9 %in% asthma_icd9 | 
      dx10 %in% asthma_icd9 | dx11 %in% asthma_icd9 | dx12 %in% asthma_icd9 | 
      dx13 %in% asthma_icd9) ~ "dx_asthma_not_primary",
    # other dx 
    !(dx1 %in% asthma_icd9 | dx2 %in% asthma_icd9 | dx3 %in% asthma_icd9 | 
      dx4 %in% asthma_icd9 | dx5 %in% asthma_icd9 | dx6 %in% asthma_icd9 | 
      dx7 %in% asthma_icd9 | dx8 %in% asthma_icd9 | dx9 %in% asthma_icd9 | 
      dx10 %in% asthma_icd9 | dx11 %in% asthma_icd9 | dx12 %in% asthma_icd9 | 
      dx13 %in% asthma_icd9) & is.na(ndc) ~ "dx_other",
    # ndc code
    (ndc %in% saba_ndc) ~ "pharm_saba",
    !is.na(ndc) & !(ndc %in% saba_ndc) ~ "pharm_other"))) %>% 
  # convert date variables to dates
  mutate_at(vars(contains("date")), funs(as.Date(.,format="%Y-%m-%d"))) %>% 
  mutate(los = as.numeric(los)) %>% 
  # join in place of service text
  left_join(place_of_service, by = "pos") %>% 
  # simpler place of service grouping
  mutate(pos_simple = as.factor(case_when(pos == "01" ~ "Pharmacy",
    pos == "03" ~ "School",
    pos == "11" ~ "Office",
    pos == "20" ~ "Urgent Care", 
    pos == "21" ~ "Inpatient Hospital",
    pos == "22" ~ "Outpatient Hospital",
    pos == "23" ~ "Emergency Room Hospital",
    pos == "41" | pos == "42" ~ "Ambulance",
    pos == "71" | pos == "72" ~ "Public Health Clinic",
    pos == "49" ~ "Independent Clinic", 
    TRUE ~ "Other")))
```

@@

I think the observations that are dx_other or pharm_other make the dataset too large and for the analyses I'm considering, may have no practical value. But they are probably worth retaining for now.

>>>>

# load outcomes vectors
load("../../data/health/outcome_list.RData")
# subset asthma and saba vector
asthma_icd9 <- pluck(outcome_icd9_list, "asthma")
saba_ndc <- pluck(outcome_icd9_list, "saba")

# read in place of service csv file
place_of_service <- read_csv("../../data/health/2013-oregon_pos.csv")

@@

```{r cohort_import}
# read path
read_path <- "../../data/health/2013-oregon_asthma_cohort.csv"
# read in asthma cohort (reading a million rows for now)
asthma_cohort <- data.table::fread(read_path, 
  # read options
  header = T, #nrows = 10000, # removing nrows read 
  colClasses = rep("character", 72)) %>% 
  # filter to oregon
  filter(STATE == "OR") %>% 
  # replace *NULL* with NA
  mutate_all(funs(replace(., .== "*NULL*" | . == "", NA))) %>% 
  # create visit type indicator 
  mutate(visit_type = as.factor(
    case_when((dx1 %in% asthma_icd9) ~ "dx_asthma_primary",
    # asthma but not as a primary
    !(dx1 %in% asthma_icd9) & (dx2 %in% asthma_icd9 | dx3 %in% asthma_icd9 | 
      dx4 %in% asthma_icd9 | dx5 %in% asthma_icd9 | dx6 %in% asthma_icd9 | 
      dx7 %in% asthma_icd9 | dx8 %in% asthma_icd9 | dx9 %in% asthma_icd9 | 
      dx10 %in% asthma_icd9 | dx11 %in% asthma_icd9 | dx12 %in% asthma_icd9 | 
      dx13 %in% asthma_icd9) ~ "dx_asthma_not_primary",
    # other dx 
    !(dx1 %in% asthma_icd9 | dx2 %in% asthma_icd9 | dx3 %in% asthma_icd9 | 
      dx4 %in% asthma_icd9 | dx5 %in% asthma_icd9 | dx6 %in% asthma_icd9 | 
      dx7 %in% asthma_icd9 | dx8 %in% asthma_icd9 | dx9 %in% asthma_icd9 | 
      dx10 %in% asthma_icd9 | dx11 %in% asthma_icd9 | dx12 %in% asthma_icd9 | 
      dx13 %in% asthma_icd9) & is.na(ndc) ~ "dx_other",
    # ndc code
    (ndc %in% saba_ndc) ~ "pharm_saba",
    !is.na(ndc) & !(ndc %in% saba_ndc) ~ "pharm_other"))) %>% 
  # convert date variables to dates
  mutate_at(vars(contains("date")), funs(as.Date(.,format="%Y-%m-%d"))) %>%
  # adding and converting some variables
  mutate(los = as.numeric(los),
         # create age variable
         age = 2013 - as.numeric(yob),
         # code in race text to numeric race
         race_factor = case_when(race == 1 ~ "White",
                                 race == 2 ~ "Black",
                                 race == 3 ~ "American Indian",
                                 race == 4 ~ "Asian",
                                 race == 5 ~ "Pacific Islander",
                                 race == 6 ~ "Other", 
                                 race == 7 ~ "Two or more",
                                 race == 9 ~ "Unknown"),
         ethnicity = case_when(ethn == "1" ~ "Hispanic",
                               ethn == "2" ~ "Not Hispanic",
                               ethn == "3" ~ "Unknown")) %>% 
  # join in place of service text
  left_join(place_of_service, by = "pos") %>% 
  # simpler place of service grouping
  mutate(pos_simple = as.factor(case_when(pos == "01" ~ "Pharmacy",
    pos == "03" ~ "School",
    pos == "11" ~ "Office",
    pos == "20" ~ "Urgent Care", 
    pos == "21" ~ "Inpatient Hospital",
    pos == "22" ~ "Outpatient Hospital",
    pos == "23" ~ "Emergency Room Hospital",
    pos == "41" | pos == "42" ~ "Ambulance",
    pos == "71" | pos == "72" ~ "Public Health Clinic",
    pos == "49" ~ "Independent Clinic", 
    TRUE ~ "Other")))
```

@@

#### Age Distribution

To understand the distribution of demographic factors, I'm going to subset the first (by date) diagnosis of asthma visit observed to look at age, gender, and race/ethnicity distributions. When you subset out the NDC and other diagnoses types, the cohort is reduced drastically. I believe this would be fine and actually create a more well-defined cohort if it's subsetted to just asthma.

Here is a count of number of persons with either a primary or other diagnosis of asthma in the cohort dataset. 
```{r age_n}
asthma_demographic <- asthma_cohort %>% 
  filter(visit_type == "dx_asthma_primary" | 
         visit_type == "dx_asthma_not_primary") %>% 
  group_by(personkey) %>% 
  arrange(todate) %>% 
  filter(row_number()==1)

nrow(asthma_demographic)
```

Summary statistics of age in persons with a medical visit related to asthma.
```{r age_sum}
summary(asthma_demographic$age)
```

Age distribution plot in persons with a medical visit related to asthma..
```{r age_dist}
# asthma age distribution plot
plot <- ggplot(asthma_demographic, aes(x=age)) +
  geom_density(color = "#19547b") +
  theme_minimal()
plot
```

#### Gender Distribution

Gender n and percentage.
```{r gender_stat}
# count
xtabs(~gender, asthma_demographic) 
# percentage
round(prop.table(xtabs(~gender, asthma_demographic)),3)*100
```

#### Race/Ethnicity Distribution

Race n and percentage. A lot of persons did not report race.
```{r race_stat}
xtabs(~race_factor, asthma_demographic)
round(prop.table(xtabs(~race_factor, asthma_demographic)),3)*100
```

Ethnicity n and percentage. Same deal. A lot of persons did not report ethnicity.
```{r ethn_stat}
xtabs(~ethnicity, asthma_demographic)
round(prop.table(xtabs(~ethnicity, asthma_demographic)),3)*100
```

### Conclusions

Overall, I think this cohort approach could work.I think the observations that are dx_other or pharm_other make the dataset too large, and may have no practical value for the analyses I'm considering. But they are probably worth retaining for now.

Some ideas are we could further limit the study population to those with at least 2 visits of asthma to further increase our specificity that a person has asthma. We could then look at the range of severity of outcomes in these people, where the most severe would be an inpatient hospital stay for a primary diagnosis of asthma. Second-most severe could be an ER/urgent care admission or ambulance visit. Least-severe could be a SABA fill in a cohort time-to-event manner. I was thinking we could start time at either the start of the wildfire period or July, and could include people who have asthma dx for some time that year, but have been without a medical bill for at least a month prior to start of study date.

For this cohort approach, I'd like to try out VanderWeele's E-value measure, which estimates the impact unmeasured confounders would have to have to explain the association.

We could also look at time-series and case-cross over study designs too like Sheryl suggested. I think this is a good idea too, and hopefully these results would agree with our general hypotheses that smoke is bad.
